---
title: "Resilience Cohort in BioHEART"
subtitle: "BioHEART"
author: "Daniel Cheng, Matthew Shu"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: lumen
    toc: yes
    toc_depth: 3
    toc_float: no
    code_folding: hide
---

```{r setup, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
```{r, results='hide', warning=FALSE, message=FALSE}
library(tidyverse)
library(dplyr)
library(glmnet)
library(caret)
library(corrplot)
library(naniar)
library(factoextra)
library(kableExtra)
library(MultiAssayExperiment)
library(boot)
library(CVrisk)
library(limma)
library(plotly)
library(Glimma)
library(ggrepel)
library(pscl)
library(bestNormalize)
library(gridExtra)
library(RiskScorescvd)
```

# Biological Age

```{r, results='hide', warning=FALSE, message=FALSE}
load("MultiAssayExperiment_20220816.RData")

## Go one folder up from working directory and into another folder 

# Mac
#data_folder = "/Users/matthewshu/Documents/GitHub-Projects/BioHEART Summer Project/Data"

# Windows
data_folder = "/Github-Projects/BioHEART-Summer-Project/Data"


BioHRT_dat_new = readxl::read_excel(file.path(data_folder, "DL5_20240829-Full-DI.xlsx"), "values")
BioHRT_dat = readxl::read_excel(file.path(data_folder,
                                         "DL4_20230831-BL_Imaging-Omics-20240116_SHU.xlsx"), "nolabel")
load("FRS_data.RData")

prs = readxl::read_excel(file.path(data_folder, "scoresStandardised-20250811.xlsx"), "scoresStandardised")
```

```{r}
head(BioHRT_dat)
BioHRT_dat = BioHRT_dat %>%
  rename(gender = sex) %>%
  left_join(BioHRT_dat_new %>% mutate(record_id = as.numeric(record_id)) %>% select(record_id, cacs_iapl_directimport), by="record_id") %>%
  mutate(cacs = coalesce(cacs, cacs_iapl_directimport)) %>%
  left_join(prs %>% dplyr::select(-bioheart_id), by="record_id")
```


## Ensemble Model of Resilience

### Calculate Scores

Relevant Variables from BioHEART

* `race` (white, aa, chinese, or hispanic) 
* `age` (years) 
* `totchol` (mg/dL) \<- ours is in mmol/L hdl (mg/dL) 
* `sbp` (mmHg) 
* `bp_med` (Y/N)
* `ace_i`
* `bblocker`
* `diur_loop`
* `diur_k`
* `diur_unk`
* `ccb smoker` (Y/N) 
* `diabetes` (Y/N)
* `bmi` 
* `lipid_med` (Y/N)
* `statin`
* `fh_heartattack` (Y/N)
* `cac` (Agatson)


#### All Score Calculations

We have decided to calculate the following CVD Risk Scores for all BioHEART subjects. There are certain limitations of each score where important re-coding decisions were made on the BioHEART data. These limitations are denoted below.

* FRS
* ASCVD (Restricted)
  + 20 < age < 79 
  + 130 < totchol < 320
  + 20 < hdl < 100
  + 90 < sbp < 200
  + White, African American, Other
* MESA CHD (Limited Ethnicities)
  + White, African American, Chinese, American
* SCORE2

BioHEART Ethnicity Coding (ethcat)

* 1 = European 
* 2 = Indigenous Australia 
* 3 = Polynesian 
* 4 = African 
* 5 = Asian 
* 6 = Indian 
* 7 = Middle Eastern
* 8 = Hispanic
* 9 = Other / Dual Ethnicity

Important Recoding Decisions
* For ACCAHA, all other ethnicities were grouped under "Other". 
* For MESA, Polynesians, Middle Eastern and any Other Ethnicities were classified as White.


```{r}

pheno_df = BioHRT_dat %>%
  filter(disc_1000 == 1) %>%
  # remove where gender or cac or cacs_pct not recorded
  drop_na(gender, cacs, cacs_pct) %>%
  # tc and hdl non-zero
  filter(!is.na(tc) | !is.na(hdl)) %>%
  # missing value encoded as .
  filter(cacs_pct != ".") %>%
  # limits of ascvd limited between 30 and 74
  #filter((age<=79) & (age>=20)) %>%
  mutate(
    # refactor gender
    gender = case_when(
      gender==1 ~ 'male',
      gender==2 ~ 'female'
      ),
    # convert from mmol/L to mg/dL
    HDL_mgdl = hdl*38.67,
    Chol_mgdl = tc*38.67,
    # new variable presence or absence of blood pressure meds
    bp_med = case_when(
      ace_arb == 1 ~ 1,
      bblocker == 1 ~ 1,
      diuretic == 1 ~ 1,
      ccb == 1 ~ 1,
      TRUE ~ 0
    ),
    # new variable presence or absence of lipid lowering meds
    lipid_med = case_when(
      statin == 1 ~ 1,
      ezetimibe == 1 ~ 1,
      fibrate == 1 ~ 1,
      niacin == 1 ~ 1,
      babr == 1 ~ 1,
      plant_sterol == 1 ~ 1,
      lipid_lowering_unknown == 1 ~ 1,
      TRUE ~ 0
    ),
    # assume no history of IHD if NA
    fh_ihd = case_when(
      fh_ihd == 1 ~ 1,
      fh_ihd == 0 ~ 0,
      is.na(fh_ihd) ~ 0
    ),
    # CACS as number
    cacs = as.numeric(cacs),
    cacs_pct = as.numeric(cacs_pct),
    # ACCAHA Ethnicity
    accaha_eth = case_when(
      ethcat == 1 ~ 'white',
      ethcat == 4 ~ 'aa',
      TRUE ~ 'other'
    ),
    # Unresolved: MESA only contains 4 ethnicity categories, what do we do with other?
    mesa_eth = case_when(
      ethcat == 1 ~ 'white',
      ethcat == 2 ~ 'aa',
      ethcat == 3 ~ 'white', # contentious Polynesian
      ethcat == 4 ~ 'aa',
      ethcat == 5 ~ 'chinese',
      ethcat == 7 ~ 'white',
      ethcat == 8 ~ 'hispanic',
      ethcat == 9 ~ 'white', # contentious mixed/Other
      TRUE ~ 'white' # if NA assume white
    ),
    
  )

pheno_df = pheno_df %>%
  rowwise() %>%
  mutate(
    # FRS Score
    ascvd_10y_frs = ascvd_10y_frs(gender, age, HDL_mgdl, Chol_mgdl, sbp, bp_med, curr_smok, cvhx_dm),
    # ASCVD (Pooled Cohort Equations - USA) w/ Requirements
    # 20 < age < 79 
    # 130 < totchol < 320
    # 20 < hdl < 100
    # 90 < sbp < 200
    ascvd_10y_accaha = ascvd_10y_accaha(accaha_eth, gender, age, Chol_mgdl, HDL_mgdl, sbp, bp_med, curr_smok, cvhx_dm),
    # MESA CHD
    chd_10y_mesa = chd_10y_mesa(mesa_eth, gender, age, Chol_mgdl, HDL_mgdl,lipid_med, sbp, bp_med, curr_smok, cvhx_dm, fh_ihd),
    # MESA CHD w/ CAC
    chd_10y_mesa_cac = chd_10y_mesa_cac(mesa_eth, gender, age, Chol_mgdl, HDL_mgdl,lipid_med, sbp, bp_med, curr_smok, cvhx_dm, fh_ihd, cacs),
    # SCORE2 <= 70 year olds
    SCORE2_new = SCORE2(Risk.region = "Low", age, gender, curr_smok, sbp, cvhx_dm, tc, hdl, FALSE),
    
    )


score_df = pheno_df %>%
  dplyr::select(record_id, cacs, cacs_pct, ascvd_10y_frs, ascvd_10y_accaha, chd_10y_mesa, SCORE2_new)
# Picked MESA without CAC since CAC will be the y-axis when calculating residuals
# 112 missing values out of range but no particular pattern in its distribution

```

Score Correlation

```{r}
# with the value labels
score_df %>%
  select(ascvd_10y_frs, ascvd_10y_accaha, chd_10y_mesa, SCORE2_new) %>%
  drop_na() %>%
  cor() %>%
  corrplot::corrplot(method = "number", type = "upper", tl.col = "black", tl.srt = 45)


```


### Calculating Zero-Inflated Regression Percentiles by Risk Score

#### Transformation and Standardisation

The following alternative is from Stuart's suggested method to standardise risk scores and take an average where possible.

```{r}

library(caret)


standardise <- function(x) {
  return((x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))

}

# Normalizing the scores with orderNorm
score_df$ordernorm_ascvd_10y_frs = standardise(orderNorm(score_df$ascvd_10y_frs)$x.t)
score_df$ordernorm_ascvd_10y_accaha = standardise(orderNorm(score_df$ascvd_10y_accaha)$x.t)
score_df$ordernorm_chd_10y_mesa = standardise(orderNorm(score_df$chd_10y_mesa)$x.t)
score_df$ordernorm_SCORE2 = standardise(orderNorm(score_df$SCORE2_new)$x.t)

hist(score_df$ordernorm_ascvd_10y_frs)
hist(score_df$ordernorm_ascvd_10y_accaha)
hist(score_df$ordernorm_chd_10y_mesa)
hist(score_df$ordernorm_SCORE2)

# Applying normalization to each score column
score_df <- score_df %>%
  rowwise() %>%
  mutate(
    average_norm_score = mean(c_across(starts_with("ordernorm_")), na.rm = TRUE)
  ) %>%
  ungroup()

score_df %>%
  # Select only columns starting with ordernorm
  select(record_id, starts_with("ordernorm_")) %>%
  filter(record_id < 100) %>%
  pivot_longer(
    cols = starts_with("ordernorm_"),
    names_to = "score",
    values_to = "value"
  ) %>%
  ggplot(aes(y = value, x = as.factor(record_id), color = score)) +
  geom_point() + 
  theme_bw()


score_df %>%
  # Select only columns starting with ordernorm
  select(record_id, starts_with("ordernorm_"), average_norm_score) %>%
  #filter(record_id < 100) %>%
  pivot_longer(
    cols = c(starts_with("ordernorm_"), average_norm_score),
    names_to = "score",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value, color = score)) +
  geom_freqpoly() +
  theme_bw()
  
```


Fits on Combined Average Normalised Score

```{r}

# Fits Zero Inflated model w/ only FRS Score

# 1 dropped because missing either CACS or average_norm_score need to investigate
score_df = score_df %>% drop_na(cacs, average_norm_score)
zeroinflmodel = zeroinfl(100*cacs ~ average_norm_score | average_norm_score, data = score_df, dist = "negbin")
summary(zeroinflmodel)

```


Fits on Traditional Risk Factors

```{r}

score_df = pheno_df %>% drop_na(cacs, sbp)
score_df %>% select(cacs, age, gender, Chol_mgdl, HDL_mgdl, sbp, bp_med, curr_smok, cvhx_dm)
zeroinflmodel = zeroinfl(100*cacs ~ age + gender + Chol_mgdl + HDL_mgdl + sbp + bp_med + curr_smok + cvhx_dm | age + gender + Chol_mgdl + HDL_mgdl + sbp + bp_med + curr_smok + cvhx_dm, data = score_df, dist = "negbin")
summary(zeroinflmodel)
```

```{r}
score_df = score_df %>% 
  mutate(
    # Predicted counts from the negative binomial part of the zero-inflated model
    # Based on the risk scores of a person, what should be their CACS
    cacs_riskscorecond_countpred = predict(zeroinflmodel, data=score_df, type="count") / 100,
    
    # Predicted probability of a zero result from the zero-inflated part of the model
    cacs_riskscorecond_zeroprob = predict(zeroinflmodel, data=score_df, type="zero"),
    
    # Overall predicted mean counts from the combined negative binomial & zero-inflated model
    cacs_riskscorecond_meanpred = predict(zeroinflmodel, data=score_df, type="response") / 100
  ) %>% 
  mutate(
    #The distribution is discrete, so for a given CACS value, I
    # will set the percentile to the mean of p(obs) <= CACS
    # and p(obs) < CACS i.e. add the two and divide by 2
    cacs_riskscorecond_pct = case_when(
      #when CACS > 0, p(obs) <= CACS and p(obs) < CACS both
      # include the probability that p(obs)==0 
      cacs > 0 ~ cacs_riskscorecond_zeroprob + (1 - cacs_riskscorecond_zeroprob) * 0.5 * (
        pnbinom(q = 100 * score_df$cacs, size = zeroinflmodel$theta, mu = 100 * cacs_riskscorecond_countpred) +
        pnbinom(q = 100 * (score_df$cacs) - 1, size = zeroinflmodel$theta, mu = 100 * cacs_riskscorecond_countpred)
      ),
      #when CACS==0, p(obs) <= CACS is just the probability
      # that p(obs)==0, because p(obs) < 0 is 0
      #Note that we can get an observation of zero either
      # because we are drawing from the zero inflation, or
      # because we are drawing from the counts model and the
      # counts model happened to generate a 0
      cacs == 0 ~ 0.5 * (cacs_riskscorecond_zeroprob + (1-cacs_riskscorecond_zeroprob)*pnbinom(q = 100 * score_df$cacs, size = zeroinflmodel$theta, mu = 100 * cacs_riskscorecond_countpred))
    )
  )

```

```{r}

options(repr.plot.width = 15, repr.plot.height = 8)

plot_theme <- theme(
  legend.key.height = unit(2.5, "cm"),
  legend.title = element_text(size = 15, angle = 90),
  legend.title.align = 0.5,
  legend.direction = "vertical",
  text = element_text(size = 20)
)

plot_guide <- guides(
  size = "none", 
  colour = guide_colourbar(title.position = "right")
)

plot_zeroinflnegbinom_truevspredcacs <- ggplot(score_df, aes(
  x = log(1 + cacs_riskscorecond_meanpred),
  y = log(1 + cacs), 
  color = cacs_riskscorecond_pct
)) + 
  geom_point(size = 3) + 
  plot_guide + 
  plot_theme

plot_zeroinflnegbinom_cacspctvscacspred <- ggplot(score_df, aes(
  x = log(1 + cacs_riskscorecond_meanpred), 
  y = cacs_riskscorecond_pct,
  color = log(1 + cacs)
)) + 
  geom_point(size = 3) + 
  plot_guide + 
  plot_theme

grid.arrange(
  plot_zeroinflnegbinom_truevspredcacs, 
  plot_zeroinflnegbinom_cacspctvscacspred, 
  ncol = 2
)


```

### Classifying by Zero-Inflated Regression Percentiles by Risk Score

```{r}

score_df$consensus_class = case_when(
  score_df$cacs_riskscorecond_pct < 0.20 ~ "resilient",
  score_df$cacs_riskscorecond_pct > 0.80 ~ "susceptible",
  ((score_df$cacs_riskscorecond_pct > 0.40) & (score_df$cacs_riskscorecond_pct < 0.60))  ~ "reference",
  TRUE ~ "ignore"
)

# Modify the factor levels in your dataframe first
score_df$consensus_class_modified <- dplyr::recode(score_df$consensus_class,
                                            "reference" = "Reference",
                                            "susceptible" = "Susceptible",
                                            "resilient" = "Resilient",
                                            "ignore" = "Other (Excluded)") 

# Plot
p = ggplot(score_df %>% filter(consensus_class != "ignore"), aes(x = age, 
                     y = log(cacs + 1), 
                     color = consensus_class_modified)) +
  geom_point() +
  scale_color_manual(values = c("Reference" = "#1F77B4",        # pastel blue
                                "Susceptible" = "#D62727",      # pastel red
                                "Resilient" = "#2ecc71")) +
  theme_bw() +
  theme(panel.grid = element_blank(),                # remove gridlines
        panel.border = element_blank(),              # remove all borders
        axis.line.x.bottom = element_line(),         # keep bottom x-axis line
        axis.line.y.left = element_line(),           # keep left y-axis line
        axis.line.x.top = element_blank(),           # remove top axis line
        axis.line.y.right = element_blank()) +       # remove right axis line
  labs(x = "Age", 
       y = "log(CACS + 1)", 
       color = "Classification")

plot(p)


# Convert to factor for plotting purposes, setting 'reference' as the reference level
score_df$consensus_class = factor(score_df$consensus_class, levels = c("reference", "resilient", "susceptible", "ignore"))

table(score_df$consensus_class)

write.csv(score_df, "resilience.csv")
```

### Comparing the Classifications
#### Metabolomics

```{r, warning=FALSE}
# Metabolomics
# Picks the first assay rlmSampleAllShort_H_batch
assays(metab_exp)
metab_exp = longFormat(bioheart_mae[,,'Metabolomics'],
                 colDataCols = c("gender","age", "smurfs", "cacs", "cacs_pct", "gensini"),
                 i = 3L)

metab_df = data.frame(metab_exp) %>%
  filter(primary != "Pool") %>%
  pivot_wider(id_cols = c(primary, colname, gender, age, smurfs, cacs, cacs_pct, gensini), 
              names_from = rowname, values_from = value) %>%
  filter(if_all(everything(), ~ .!=".")) %>%
  transform(cacs = as.numeric(cacs), cacs_pct = as.numeric(cacs_pct), gensini = as.numeric(gensini), primary = as.numeric(primary)) %>%
  # Remove duplicate rows, .keep_all <- keeps all of the rest of columns
  distinct(primary, .keep_all = TRUE) %>%
  drop_na() #%>%
  #mutate(across(8:60, ~(.-min(.))/((max(.)-min(.)))))

graph_df = score_df %>%
  dplyr::select(record_id, consensus_class) %>%
  inner_join(metab_df, by=c('record_id'='primary')) # %>%
  #filter(consensus_class %in% c("resilient", "reference"))

# Graph Box Plot
graph_df %>%
  pivot_longer(cols = -c(record_id, gender, consensus_class, colname, age, smurfs, cacs, cacs_pct, gensini), names_to = "metabolite", values_to = "value") %>%
  ggplot() +
  aes(x = metabolite, y = value, colour = consensus_class) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

##### Differential Expression Analysis

The contrasts are `resilient` - `reference`.

```{r, warning=FALSE}
filtered_graph_df = graph_df #%>%
  #filter(gender == 1) %>%
  #filter(age > 60) 
lm_df = filtered_graph_df %>%
  dplyr::select(-c(record_id, consensus_class, colname, gender, age, smurfs, cacs, cacs_pct, gensini)) %>%
  t()

design = model.matrix(~ consensus_class, filtered_graph_df)
fit = lmFit(lm_df, design)
efit = eBayes(fit)
topTable(efit)
```

The contrasts are `resilient` - `reference`.

```{r, warning=FALSE}

# Contrasts
CM = makeContrasts(consensus_classresilient = consensus_classresilient, levels = design)
constrast_fit = contrasts.fit(efit, contrast = CM)
constrast_fit = eBayes(constrast_fit)

topTable(constrast_fit, n=50)

```

The contrasts are `susceptible` - `reference`.

```{r, warning=FALSE}

# Contrasts

CM = makeContrasts(consensus_classsusceptible = consensus_classsusceptible, levels = design)
constrast_fit = contrasts.fit(efit, contrast = CM)
constrast_fit = eBayes(constrast_fit)

topTable(constrast_fit, n=50)

```


The contrasts are `susceptible` - `resilient`.

```{r, warning=FALSE}

# Contrasts

CM = makeContrasts(consensus_classcompare = consensus_classsusceptible - consensus_classresilient, levels = design)
constrast_fit = contrasts.fit(efit, contrast = CM)
constrast_fit = eBayes(constrast_fit)

topTable(constrast_fit, n=50)

```


```{r, warning=FALSE}
# Volcano Plot
library(EnhancedVolcano)
p = EnhancedVolcano(
  constrast_fit,
  lab = rownames(constrast_fit),
  x = "coefficients",
  y = "p.value",
  title = "Resilient vs Reference (un-adjusted p-values)",
  pCutoff = 0.05,
  labSize = 6
)

ggsave("resilient_reference_volcano.png", plot = p, width = 12, height = 8, dpi = 300)

# MA Plot
glMDPlot(constrast_fit,
         counts = lm_df,
         coef=1, 
         main= "Resilient v.s. Reference",
         groups = filtered_graph_df$consensus_class,
         html="resilient_reference_ma_plot.html")

```

#### Lipidomics
Lipid Totals.

```{r, warning=FALSE}
lipid_totals_exp = longFormat(bioheart_mae[,,'Lipidomics_totals'],
                 colDataCols = c("age", "smurfs", "cacs", "cacs_pct", "gensini"),
                 i = 2L)

lipid_totals_df = data.frame(lipid_totals_exp) %>%
  filter(primary != "Pool") %>%
  pivot_wider(id_cols = c(primary, colname, age, smurfs, cacs, cacs_pct, gensini), 
              names_from = rowname, values_from = value) %>%
  transform(cacs = as.numeric(cacs), cacs_pct = as.numeric(cacs_pct), gensini = as.numeric(gensini), primary = as.numeric(primary))

graph_df = score_df %>%
  select(record_id, consensus_class) %>%
  inner_join(lipid_totals_df, by=c('record_id'='primary'))

score_df %>%
  select(record_id, consensus_class) %>%
  inner_join(lipid_totals_df, by=c('record_id'='primary')) %>%
  filter(consensus_class %in% c("resilient", "reference")) %>%
  pivot_longer(cols = -c(record_id, consensus_class, colname, age, smurfs, cacs, cacs_pct, gensini), names_to = "lipid_total", values_to = "value") %>%
  ggplot() +
  aes(x = lipid_total, y = value, colour = consensus_class) +
  geom_boxplot() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

##### Differential Expression Analysis

```{r, warning=FALSE}
filtered_graph_df = graph_df #%>%
  #filter(gender == 1) %>%
  #filter(age > 60) 
lm_df = filtered_graph_df %>%
  dplyr::select(-c(record_id, consensus_class, colname, age, smurfs, cacs, cacs_pct, gensini)) %>%
  t()

design = model.matrix(~ consensus_class, filtered_graph_df)
fit = lmFit(lm_df, design)
efit = eBayes(fit)
topTable(efit)

```

The contrasts are `resilient` - `reference`.

```{r, warning=FALSE}

# Contrasts
CM = makeContrasts(consensus_classresilient = consensus_classresilient, levels = design)
constrast_fit = contrasts.fit(efit, contrast = CM)
constrast_fit = eBayes(constrast_fit)

topTable(constrast_fit, n=50)

```

The contrasts are `susceptible` - `reference`.

```{r, warning=FALSE}

# Contrasts

CM = makeContrasts(consensus_classsusceptible = consensus_classsusceptible, levels = design)
constrast_fit = contrasts.fit(efit, contrast = CM)
constrast_fit = eBayes(constrast_fit)

topTable(constrast_fit, n=50)

```


The contrasts are `susceptible` - `resilient`.

```{r, warning=FALSE}

# Contrasts

CM = makeContrasts(consensus_classcompare = consensus_classsusceptible - consensus_classresilient, levels = design)
constrast_fit = contrasts.fit(efit, contrast = CM)
constrast_fit = eBayes(constrast_fit)

topTable(constrast_fit, n=50)

```


#### Proteomics

```{r, warning=FALSE}

prot_exp = longFormat(bioheart_mae[,,'Proteomics'],
                      i = "loess_balanced",
                      colDataCols = c("age", "gender", "smurfs", "cacs", "cacs_pct", "gensini"))



prot_df = data.frame(prot_exp, check.names=FALSE) %>%
  filter(!grepl("Repeat", colname)) %>%
  filter(primary != "79") %>%
  filter(primary != "Pool") %>%
  filter(primary != "Pool") %>%
  pivot_wider(id_cols = c(primary, age, gender, smurfs, cacs, cacs_pct, gensini), 
              names_from = rowname, values_from = value) %>%
  mutate(
    cacs = as.numeric(cacs),
    cacs_pct = as.numeric(cacs_pct),
    gensini = as.numeric(gensini),
    primary = as.numeric(primary)
  )

paste(colnames(prot_df), collapse=", ")

write.csv(prot_df, "proteomics.csv")
```


##### Differential Expression Analysis

```{r, warning=FALSE}

prot_patient_df = score_df %>%
  dplyr::select(record_id, consensus_class) %>%
  # Remove outlier 1067
  #filter(record_id != 1067) %>%
  dplyr::inner_join(prot_df, by=c('record_id'='primary'))
setdiff(score_df$record_id, prot_df$primary)

lm_df = prot_patient_df %>%
  select(-c(record_id, consensus_class, age, gender, smurfs, cacs, cacs_pct, gensini)) %>%
  t()

design = model.matrix(~ consensus_class, prot_patient_df)

fit = lmFit(lm_df, design)
efit = eBayes(fit)
topTable(efit)

```

The contrasts are `resilient` - `reference`.

```{r, warning=FALSE}

# Contrasts
CM = makeContrasts(consensus_classresilient = consensus_classresilient, levels = design)
constrast_fit = contrasts.fit(efit, contrast = CM)
constrast_fit = eBayes(constrast_fit)

topTable(constrast_fit, n=50)

```

The contrasts are `susceptible` - `reference`.

```{r, warning=FALSE}

# Contrasts

CM = makeContrasts(consensus_classsusceptible = consensus_classsusceptible, levels = design)
constrast_fit = contrasts.fit(efit, contrast = CM)
constrast_fit = eBayes(constrast_fit)

topTable(constrast_fit, n=50)

```


The contrasts are `susceptible` - `resilient`.

```{r, warning=FALSE}

# Contrasts

CM = makeContrasts(consensus_classcompare = consensus_classsusceptible - consensus_classresilient, levels = design)
constrast_fit = contrasts.fit(efit, contrast = CM)
constrast_fit = eBayes(constrast_fit)

topTable(constrast_fit, n=50)

```


```{r}
signif_igs = c("IGLV7.43.IGLV7.46", "IGHM", "IGKV2.29.IGKV2.30.IGKV2D.29", "IGKV2.29.IGKV2D.29")

#signif_igs = c("IGLV7-43;IGLV7-46", "IGHM", "IGKV2-29;IGKV2-30;IGKV2D-29", "IGKV2-29;IGKV2D-29")

# Immunoglobulins vs CACS Percentile

prot_patient_df %>%
  select(cacs_pct, signif_igs, consensus_class, average_norm_score) %>%
  pivot_longer(names_to = "protein", values_to = "value", -c(consensus_class, cacs_pct, average_norm_score)) %>%
  filter(consensus_class == "resilient") %>%
  ggplot() +
  aes(x = cacs_pct, y = value, colour = protein) +
  geom_point() +
  theme_bw() +
  labs(title = "Immunoglobulins vs CACS Percentile", x = "CACS Percentile", y = "Immunoglobulins")


prot_patient_df %>%
  select(cacs_pct, `IGLV7.43.IGLV7.46`, consensus_class, average_norm_score) %>%
  filter(consensus_class == "resilient") %>%
  ggplot() +
  aes(x = average_norm_score, y = `IGLV7.43.IGLV7.46`) +
  geom_point() +
  theme_bw() +
  labs(title = "Immunoglobulins vs Average Normalised Score", x = "Average Normalised Score", y = "Immunoglobulins")

# Immunoglobulins vs Average Normalised Score

prot_patient_df %>%
  select(cacs_pct, signif_igs, consensus_class, average_norm_score) %>%
  pivot_longer(names_to = "protein", values_to = "value", -c(consensus_class, cacs_pct, average_norm_score)) %>%
  filter(consensus_class == "resilient") %>%
  ggplot() +
  aes(x = average_norm_score, y = value, color = protein) +
  geom_point() +
  theme_bw() +
  labs(title = "Immunoglobulins vs Average Normalised Score", x = "Average Normalised Score", y = "Immunoglobulins")


prot_patient_df %>%
  select(cacs_pct, signif_igs, consensus_class, average_norm_score) %>%
  pivot_longer(names_to = "protein", values_to = "value", -c(consensus_class, cacs_pct, average_norm_score)) %>%
  filter(consensus_class %in% c("resilient", "susceptible")) %>%
  ggplot() +
  aes(x = average_norm_score, y = value, color = consensus_class, shape = protein) +
  geom_point() +
  theme_bw() +
  labs(title = "Immunoglobulins vs Average Normalised Score", x = "Average Normalised Score", y = "Immunoglobulins")


is_outlier <- function(x) {
  return(x < quantile(x, 0.25) - 1.5 * IQR(x) | x > quantile(x, 0.75) + 1.5 * IQR(x))
}

prot_patient_df %>%
  select(record_id, cacs_pct, signif_igs, consensus_class, average_norm_score) %>%
  filter(consensus_class %in% c("resilient", "reference", "susceptible")) %>%
  pivot_longer(names_to = "protein", values_to = "value", -c(record_id, consensus_class, cacs_pct, average_norm_score)) %>%
  mutate(outlier = ifelse(is_outlier(value), record_id, as.numeric(NA))) %>%
  ggplot() +
  aes(x = consensus_class, y = value, color = consensus_class) +
  facet_wrap(~protein)+
  geom_boxplot(notch=TRUE) +
  geom_text(aes(label = outlier), na.rm = TRUE, hjust = -0.3) +
  theme_bw() +
  labs(title = "Immunoglobulins vs Average Normalised Score", x = "Average Normalised Score", y = "Immunoglobulins")


# with the value labels
prot_patient_df %>%
  select(signif_igs) %>%
  cor() %>%
  corrplot::corrplot(method = "number", type = "upper", tl.col = "black", tl.srt = 45)

# Immunoglobulins vs ln(CACS+1)

prot_patient_df %>%
  select(cacs, signif_igs, consensus_class, average_norm_score) %>%
  pivot_longer(names_to = "protein", values_to = "value", -c(consensus_class, cacs, average_norm_score)) %>%
  filter(consensus_class %in% c("resilient", "susceptible")) %>%
  ggplot() +
  aes(x = log(cacs+1), y = value, colour = consensus_class) +
  facet_wrap(~protein, scales = "free") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(title = "Immunoglobulins vs ln(CACS+1)", x = "ln(CACS+1)", y = "Immunoglobulins")

library(ggpubr)
library(ggthemes)

# Immunoglobulins vs ln(CACS+1)

prot_patient_df %>%
  select(cacs, signif_igs, consensus_class, average_norm_score) %>%
  pivot_longer(names_to = "protein", values_to = "value", -c(consensus_class, cacs, average_norm_score)) %>%
  filter(consensus_class %in% c("resilient", "susceptible")) %>%
  mutate(ln_cacs = log(cacs+1)) %>%
  ggscatter(
    x = "ln_cacs", y = "value",
    color = "consensus_class", palette = "jco",
    cor.coef = TRUE
    ) +
    geom_smooth(method = "lm", color = "black") +
    facet_wrap(~protein, scales = "free") +
    labs(title = "Immunoglobulins vs ln(CACS+1)", x = "ln(CACS+1)", y = "Immunoglobulins")

# Immunoglobulins vs Average Normalised Score

prot_patient_df %>%
  select(cacs_pct, signif_igs, consensus_class, average_norm_score) %>%
  pivot_longer(names_to = "protein", values_to = "value", -c(consensus_class, cacs_pct, average_norm_score)) %>%
  filter(consensus_class %in% c("resilient", "susceptible")) %>%
  ggscatter(
    x = "average_norm_score", y = "value",
    color = "consensus_class", palette = "jco",
    cor.coef = TRUE
    ) +
    geom_smooth(method = "lm", color = "black") +
    facet_wrap(~protein, scales = "free") +
    labs(title = "Immunoglobulins vs Average Normalised Score", x = "Average Normalised Score", y = "Immunoglobulins")
```

